Mapping Cancer: the Potential of Cartograms and Alternative Map Displays

### 2020-05-22 literature.Rmd
revisions by di

Epidemiologists and statisticians have developed the statistics used to communicate the burden of cancer over several decades. Table 1<!--\ref{tab:measures}--> summarizes the measures commonly presented in published cancer atlases. Mortality rates are commonly presented as relative rates of risk across the population and age-adjusted to correct for the higher prevalence of cancers in older populations. As described in Howe [@HEDP], Englishman P. Stocks advanced the field of mortality statistics by introducing the standardized mortality ratios in the 1930s, which is an improvement on crude death rates.

XXX
Some of the measures used to present cancer statistics are aggregations. The counts of cases in small areas can be difficult to obtain as they are often protected for privacy reasons. The information released and the statistics presented in worldwide cancer atlases are often modelled estimates. The measures described in table use Incidence as the statistic of interest, this is the amount of new cases and the same statistics can be used to describe the mortality, in this case Standardised Incidence Rate is analogous to the Standarised Mortality Rate.

@@

Figure 4<!--\ref{fig:auscartograms}--> shows four main types of cartograms using melanoma incidence on Australian Statistical Areas at Level 3 [@abs2016]. The version of a contiguous cartogram (a) has expanded the highly populated areas while preserving the full shapes of rural areas. 
The south east is enlarged, but high population areas are still small, and low population areas are still large on the map. It has not fully resolved the population transformation of areas, the algorithm doesn't reach an optimal configuration where area matches population -- Australia is too heterogeneous for the algorithm to handle. The shape-preserved cartogram is unreadable, and it has reduced all areas to tiny spots on the map. Zooming in on a high-resolution output shows it does preserve the shapes. The Dorling cartogram and the hexagon tile map provide reasonable displays of the spatial distribution, despite having a very large amount of white space in the outback areas.

@@

Cartograms provide the spatial distortion to more accurately convey the statistical distribution, focusing on the human impact of the disease. However, the transformation of contiguous cartograms often occurs at the expense of the shape of areas [@CBATCC, @NAC, @TAAM]. When the population density of the geographic units is highly dissonant with geographic density, the cartogram will lose all spatial context. Dorling [@ACTUC] has a cartogram showing the 1966 general election results, which looked very little like the geographical shape of Australia. XXX

@@

Table 3: <!--\label{tab:methods}--> Summary of features and constraints of common mapping methods used to display cancer statistics. Yes means that the map dispay always employs the characteristic. No means that the display does not employ the characteristic. Possible means it is possible to employ the characteristic when using the display, but it is not necessary to produce the display.

@@

Choropleth map displays are familiar and useful for map users to orient themselves to the statistic and additional features of the display. A choropleth map display may be the most effective display for the data considering budget, time, and maintenance contraints. Map creators should consider the shapes and discrepancies in size within the set of geographic units used for analysis, and whether the features of the dataset warrant transformation of the geographic units.
Additional mapping methods should be considered by map creators during the development of a cancer atlas, as alternative displays may align better with the purpose of the display. Incorporating the population of the geographic units may better inform the map readers of the distribution of the statistic. The importance and potential impact of utilising a display that emphasises the underlying population should be considered during the design phase.
Cartograms and tile maps should be adopted as primary displays going forward, or secondary displays if the familiarity of the choropleth map will benefit the users. Clear explanations of the interpretation of the geogephic units in alternative displays should also be included, along with the features of the distribution of the statistic.

>>>>

Epidemiologists and statisticians have developed a range of statistics to communicate the burden of cancer, and the choice of statistics used in maps has changed in recent decades. Table 1<!--\ref{tab:measures}--> summarizes the measures commonly presented in published cancer atlases. Mortality rates are commonly presented as relative rates of risk across the population, and age-adjusted to correct for the higher prevalence of cancers in older populations. As described by Howe [@HEDP], Englishman P. Stocks advanced the field of mortality statistics by introducing the standardized mortality ratios in the 1930s, which was an improvement on crude death rates.

The measures displayed are typically aggregations of data values over small areas or model estimates. This to protect patient privacy, and for numerical stability. The counts of cases in small areas can be difficult to obtain as they are often protected for privacy reasons. The information released and the statistics presented in worldwide cancer atlases are often model estimates. The measures described in Table 1 use incidence as the statistic of interest, which reports the number of new cases. These same statistics can be used to describe the mortality, that is, Standardised Incidence Rate is analogous to the Standarised Mortality Rate.

@@

Figure 4<!--\ref{fig:auscartograms}--> shows four main types of cartograms using melanoma incidence on Australian Statistical Areas at Level 3 [@abs2016]. The version of a contiguous cartogram (a) has expanded the highly populated areas while preserving the full shapes of rural areas. 
The south east is enlarged, but high population areas are still small, and low population areas are still large on the map. It has not fully resolved the population transformation of areas, because the algorithm can't reach an optimal configuration where area matches population -- Australia is too heterogeneous. The shape-preserved cartogram is unreadable, and it has reduced all areas to tiny spots on the map. Zooming in on a high-resolution output shows it does preserve the shapes. The Dorling cartogram and the hexagon tile map provide reasonable displays of the spatial distribution, despite having a very large amount of white space in the outback areas.

@@

Cartograms provide the spatial distortion to more accurately convey the statistical distribution, focusing on the human impact of the disease. However, the transformation of contiguous cartograms often occurs at the expense of the shape of areas [@CBATCC, @NAC, @TAAM]. When the population density of the geographic units is highly dissonant with geographic density, the cartogram will lose all spatial context. Dorling [@ACTUC] contains a cartogram showing the 1966 general election results for Australia, produced Hughes and Savage [@HughesSavage], which looked very little like the geographical shape of Australia. The reader is encouraged to access the freely available pdf of Dorling's book, and this image can be found on page 41.

@@

Table 3: <!--\label{tab:methods}--> Summary of features and constraints of common mapping methods used to display cancer statistics. ("Yes" indicates the characteristic is always used, "No" indicates never used, and "Possible" it may be employed, but it is not necessary.)

@@

<!-- recommendation -->
Additional mapping methods should be considered by map creators during the development of a cancer atlas, as alternative displays may align better with the communication purpose. Other considerations need to be taken into account, including audience, budget, time, maintenance. Of primary importance is that information about cancer statistics be effectively communicated to the public. The choropleth has an advantage in that it is more familiar in form for more people, but we have seen that it can give an incorrect perception of the information. Public atlases can be useful educational tools. A combination of the choropleth alongside an alternative display, could provide a balance of the familiar along with a perceptually accurate display, and provide an opportunity to educate the reader.  This is especially recommended for Australia, because of the vast difference between spatial area and population density.
